Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-14 paper/preprint/paper.tex
just rerendered

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
showed that artificial neural networks (ANN) are sensitive to
adversarial examples: perturbed versions of data instances that yield
vastly different model predictions despite being ``imperceptible'' in
that they are semantically indifferent from their factual counterparts.
Even though some partially effective mitigation strategies have been
proposed---most notably \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
Szegedy 2014})---truly robust deep learning (DL) remains unattainable
even for models that are considered shallow by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so-called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}) and generalizability
(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance
alone, the solutions may seem to provide compelling explanations for the
data, when in fact they are based on purely associative, semantically
meaningless patterns. This poses two related challenges. Firstly, there
is no dependable way to verify if such complex representations
correspond to meaningful and plausible explanations. Secondly, even if
we could resolve the first challenge, it remains undecided how to ensure
that models can \emph{only} learn valuable explanations.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements; the former means that
the counterfactuals should comply with the distribution of the factual
data, the latter means that they should respect arbitrary (global)
actionability constraints. To make the desiderata for our framework more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that CEs---understood here
as minimal input perturbations that yield some desired model
prediction---are generally more meaningful if the underlying model is
more robust to adversarial examples. We can make intuitive sense of this
finding when looking at adversarial training (AT) through the lens of
representation learning with high degrees of freedom. As argued before,
learned representations may be sensitive to producing implausible
explanations and mispredicting for worst-case counterfactuals (i.e.,
adversarial examples). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CEs and AEs are generated,
Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
case for jointly studying the opaqueness and robustness problems in
representation learning. Formally, AEs can be seen as the subset of CEs
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CEs and AEs are equivalent under certain conditions and derive
theoretical upper bounds on distances between them.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in detail
below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
involved in \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and
Mordatch 2020}). The trade-off between the components can be governed by
adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

We run the experiments with three gradient-based generators:
\emph{Generic} of Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
approach, \emph{REVISE} (\citeproc{ref-joshi2019realistic}{Joshi et al.
2019}) that aims to generate plausible counterfactuals using a surrogate
Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023}) but without
the conformal prediction component---as a method that directly targets
both faithfulness and plausibility of the CEs.

@@

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter major
variability in the outcomes. CT is exposed to two potential sources of
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
recognize this to be a challenge for \emph{ECCCo} and so it may have
downstream impacts on our proposed method. Still, we find that training
instabilities can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating a sufficiently large number of
counterfactuals during each training epoch, and including only mature
counterfactuals for contrastive divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from tuning certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian (e.g.,
\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
al. 2017}) optimization.

\subsection{Limitations and Challenges}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs\footnote{See supplementary appendix for computational details.}.
However, there are three factors that attenuate the impact of this
limitation. First, CT provides counterfactual explanations for the
training samples essentially for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even with sufficient domain knowledge to protect all
immutable features, there may exist proxies that are theoretically
mutable (and hence should not be protected) but preserve enough
information about the principals to hinder the protections. As an
example, consider the Adult dataset used in our experiments where the
mutable education status is a proxy for the immutable age, in that the
attainment of degrees is correlated with age. Delineating actionability
is a major undecided challenge in the AR literature (see, e.g.,
\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
and Alfano 2020}) impacting the capacity of CT to increase the
explainability of the model.

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, CT could result in an unfairly assigned burden of
recourse (e.g., \citeproc{ref-sharma2020certifai}{Sharma, Henderson, and
Ghosh 2020}), threatening the equality of opportunity in the system
(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
Lakkaraju 2023}). Still, as the referenced publications indicate, such
phenomena are not specific to CT; all types of ADM solutions without
strong external protections have been recognized to promote harmful
power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).

@@

\bibitem[\citeproctext]{ref-altmeyer2023faithful}
Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
S. Liem. 2023. {``Faithful Model Explanations Through Energy-Constrained
Conformal Counterfactuals.''} \url{https://arxiv.org/abs/2312.10648}.

@@

\bibitem[\citeproctext]{ref-gao2023impact}
Gao, Ruijiang, and Himabindu Lakkaraju. 2023. {``On the Impact of
Algorithmic Recourse on Social Segregation.''} In \emph{Proceedings of
the 40th International Conference on Machine Learning}. ICML'23.
Honolulu, Hawaii, USA: JMLR.org.

@@

\bibitem[\citeproctext]{ref-maas2023machine}
Maas, Jonne. 2023. {``Machine Learning and Power Relations.''} \emph{AI
\& SOCIETY} 38 (4): 1493--1500.

@@

\bibitem[\citeproctext]{ref-morcos2019success}
Morcos, Ari S., Haonan Yu, Michela Paganini, and Yuandong Tian. 2019.
{``One Ticket to Win Them All: Generalizing Lottery Ticket
Initializations Across Datasets and Optimizers.''} In \emph{Proceedings
of the 33rd International Conference on Neural Information Processing
Systems}. Red Hook, NY, USA: Curran Associates Inc.

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Package Version (Reproducibility)}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

>>>>

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}) showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
Szegedy 2014})---truly robust deep learning remains unattainable even
for models that are considered ``shallow'' by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so-called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}). Similarly, Zhang et al.
(\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are expressive enough to fit randomly
labeled data. Thus, looking at the predictive performance alone, the
solutions may seem to provide compelling explanations for the data, when
in fact they are based on purely associative, semantically meaningless
patterns. This poses two challenges. Firstly, there is no dependable way
to verify if representations correspond to meaningful, plausible
explanations. Secondly, even if we could resolve the first challenge, it
remains undecided how to ensure that models can \emph{only} learn
valuable explanations.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and (global) actionability constraints. To make the
desiderata for our framework more concrete, we follow Augustin, Meinke,
and Hein (\citeproc{ref-augustin2020adversarial}{2020}) in tying the
concept of explainability to the quality of CEs that can be generated
for a given model. The authors show that CEs---understood as minimal
input perturbations that yield some desired model prediction---tend to
be more meaningful if the underlying model is more robust to adversarial
examples. We can make intuitive sense of this finding when looking at
adversarial training (AT) through the lens of representation learning
with high degrees of freedom. As argued before, learned representations
may be sensitive to producing implausible explanations and mispredicting
for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
``unlearn'' susceptiblity to such examples, AT can effectively remove
implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated, Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) makes the case for
jointly studying the opaqueness and robustness problems in
representation learning. Formally, AEs can be seen as the subset of CEs
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CEs and AEs are equivalent under certain conditions and derive
theoretical upper bounds on distances between them.

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
between the components can governed through penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

We run the experiments with three gradient-based generators:
\emph{Generic} of Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
approach, \emph{REVISE} (\citeproc{ref-joshi2019realistic}{Joshi et al.
2019}) that aims to generate plausible counterfactuals using a surrogate
Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
Altmeyer et al. (\citeproc{ref-altmeyer2024faithful}{2024}) but without
the conformal prediction component---as a method that directly targets
both faithfulness and plausibility of the CEs.

@@

\textbf{\emph{CT is subject to training instabilities.}} JEMs are
susceptible to instabilities during training
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter major
variability in the outcomes. CT is exposed to two potential sources of
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
Still, we find that training instabilities can be successfully mitigated
by regularizing energy (\(\lambda_{\text{reg}}\)), generating
sufficiently many counterfactuals during each training epoch, and
including only mature counterfactuals for contrastive divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from the tuning of certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian (e.g.,
\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
al. 2017}) optimization.

\subsection{Current Limitations}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature (see, e.g.,
\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
and Alfano 2020}) impacting the capacity of CT to increase the
explainability of the model.

\textbf{\emph{Interventions on features may impact fairness.}} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features, which may have implication for
the fair and equitable treatment of decision subjects. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse (e.g.,
\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
threatening the equality of opportunity in the system
(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Still, these
phenomena are not specific to CT.

@@

@@

@@

@@

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Package Version (Reproducibility)}, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, colframe=quarto-callout-warning-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
